Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-29 paper-1/manuscript.tex
Editing

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Repetition suppression indices $<$msc - std$>$ across all networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics. \textbf{c} Deviance detection indices $<$dev - msc$>$ across all networks and stimuli

Having confirmed the presence of deviance detection in our model, we then turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. In the remainder of the paper, we will proceed to detail the processes leading to deviance detection on a sample network, then confirm the highlighted trends statistically with reference to the full set of 120 networks and stimuli. The sample network was chosen to be roughly representative of the population as follows: We assessed each network's average response to stimulation with A and B in the msc sequence and calculated z-scores with respect to the population. We then focused on networks whose absolute z-score was less than 1 for both A and B responses, and hand-picked a sample that exhibited both an increased DDI in STD+TA over TA only, and positive RSI in both STD+TA and TA only conditions.

@@

In the following, we will use a sample network and target stimulus to first build an understanding of how TA leads to deviance detection when acting alone, i.e. in the ablated TA only model. Then, we will build on this understanding to elucidate how the addition of STD, which was shown above to be ineffective for deviance detection on its own, can lead to an increased deviant response in the full model.

@@

To understand why the same stimulus A evoked a higher response in dev than in msc with TA only, we first compared the average responses in the two conditions, sorting neurons by their response onset to permit a high-resolution view of the population activity. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both conditions. The contrast (right) revealed a tendency of dev spikes to occur earlier, particularly shortly after stimulation, as shown by the succession of orange (higher dev activity) by blue (higher msc activity) in the lower parts of the panel. In later parts of the response (top half of the panel), the dev response was larger, but not obviously shifted in time. Since the stimulus was identical in both conditions, and TA was the only plsaticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

We quantified the former with the strength of adaptation $V_{TA}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target dev trials relative to msc, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). In contrast, once the bulk of activity had occurred, many neurons showed higher thresholds in dev (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $V_{TA}$.

To assess the difference in synaptic currents in a way that allowed direct comparison to $V_{TA}$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs (resets after spikes notwithstanding). As the direct cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in dev around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

@@

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800), and this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Median $V_{TA}$ was lower in dev than msc (T = 514, p = 1.67e-16, median = -0.094 mV; one-sided Wilcoxon, n = 120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two conditions. In order to assess the relative contribution of $V_{TA}$ to the increased dev response, we chose to focus exclusively on time points with higher firing probability in the dev sequence, discarding data points with equal or lower firing probability relative to the msc sequence. We then weighted the remaining contrasts in $V_{TA}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased dev response. While thresholds were slighly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $V_{TA}$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $V_{TA}$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing in the beginning, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

First, we mapped the average value of $V_{TA}$ at the start of target trials to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the msc sequence, $V_{TA}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the dev sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.
Next, we turned to the likely cause of this difference, the non-target responses (Figure \ref{fig5}b), mapped into space as the average number of spikes fired in each neuron. As expected, the pattern of activity in non-target trials, constituting 80\% of all trials, closely matched that of the thresholds. Notably, although the comparison is between 400 B trials in dev, and only 100 B trials (alongside 300 C, D and E trials) in msc, only a small handful of neurons responded more to non-target stimulation in dev, most of these very close to the stimulation site of B.

Correlating the non-target response contrast (dev - msc) with the contrast in $V_{TA}$ across excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response between msc and dev sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
% TODO 5-8: Re-label stim locations once per figure
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any condition. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (dev - msc) and the contrast in $V_{TA}$ at the start of B trials, and associated linear regression. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are plotted with crosses, while later responders are plotted with dots. Inset histograms show the distributions of the corresponding values across these two categories, and the confidence intervals of their medians, calculated by bootstrap with 10000 iterations.
\textbf{c} Median non-target response contrast (dev - msc) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labelled "late", across networks and stimuli.
\textbf{c} Median contrast (dev - msc) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and thereby decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (a) responds very quickly to stimulation in B, (b) responds more strongly to non-target stimuli in dev than msc, and (c) has increased TA levels as a result, reducing the later response in the same way as demonstrated in Figure \ref{fig4} in reverse.

To see this in the sample network, consider first the latency of responses to B across all conditions, shown by way of rank order in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward from the stimulus site. We did notice that the early response appears to gravitate towards the stimulus site for A, but confirmed that this was the case in all conditions separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B in any condition, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the response contrast (dev - msc) to the contrast in $V_{TA}$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993) echoed the related finding in Figure \ref{fig5}c. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the dev sequence, while the response of the network as a whole decreased. Correspondingly, the TA levels of this portion of the network also tended to increase, in clear contrast to the network as a whole.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $V_{TA}$ contrasts were robustly correlated (Median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon test for $\rho > 0$, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $V_{TA}$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in dev than msc (T = 6.98e+03, p = 8.48e-19; one-sided Wilcoxon, n = 120), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $V_{TA}$ followed the same pattern, with the early portion exhibiting stronger TA in dev (T = 4.63e+03, p = 2.21e-17), while the late portion was more recovered (T = 26, p = 2.97e-17).

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. Lines show the average (trial mean +- SEM, left axis) network-wide spike counts in target trials in each sequence. Boxes summarize the spike count totals per trial (right axis), the means of which form the basis for index calculations (see panel c). Note the different axis scaling on the top row (without TA) and the bottom row (with TA). \textbf{b} Repetition suppression indices $<$msc - std$>$ across networks and stimuli. Asterisks indicate data greater than 0, established with a Wilcoxon signed-rank test at a significance level 0.05. See main text for detailed statistics. \textbf{c} Deviance detection indices $<$dev - msc$>$ across networks and stimuli

Having confirmed the presence of deviance detection in our model, we turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. In the remainder of the paper, we will proceed to detail the processes leading to deviance detection on a sample network, then confirm the highlighted trends statistically with reference to the full set of 120 networks and stimulus assignments. The sample network was chosen to be roughly representative of the population as follows: We assessed each network's average response to stimulation with A and B in the msc sequence and calculated z-scores with respect to the population. Considering only networks whose absolute z-score was less than 1 for both A and B responses, we hand-picked a sample network and stimulus assignment that exhibited both an increased DDI in STD+TA over TA only, and positive RSI in both STD+TA and TA only conditions.

@@

In the following, we will first build an understanding of how TA leads to deviance detection when acting alone, i.e. in the ablated TA only model, working backwards from the increased target response in dev, identifying and tracing causal factors step by step. Then, we will build on this understanding to elucidate how the addition of STD, which was shown above to be ineffective for deviance detection on its own, can lead to an increased deviant response in the full model.

@@

To understand why the same stimulus A evoked a higher response in dev than in msc with TA only, we first compared the average responses in the two conditions, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both conditions. The contrast (right) revealed a tendency of dev spikes to occur earlier, particularly shortly after stimulation, as shown by the succession of orange (higher dev activity) by blue (higher msc activity) in the lower parts of the panel. In later parts of the response (top half of the panel), the dev response was larger, but not obviously shifted in time. Since the stimulus was identical in both conditions, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

We quantified the former with the strength of adaptation $V_{TA}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target dev trials relative to msc, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in dev (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $V_{TA}$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $V_{TA}$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in dev around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

@@

Based on these measures, we then asked why there were more spikes in dev trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in dev, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response of each excitatory neuron in terms of spikes fired. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in dev than msc (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: Median $V_{TA}$ was lower in dev than msc (T = 514, p = 1.67e-16, median = -0.094 mV; one-sided Wilcoxon, n = 120), and the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that networks were less adapted during oddball sequences, and that this lack of adaptation played an important role in the increased response to deviant inputs.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two conditions. In order to assess the relative contribution of $V_{TA}$ to the increased dev response, we chose to focus exclusively on time points with higher firing probability in the dev sequence, discarding data points with equal or lower firing probability relative to the msc sequence. We then weighted the remaining contrasts in $V_{TA}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased dev response. While thresholds were slighly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $V_{TA}$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $V_{TA}$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

First, we mapped the average value of $V_{TA}$ at the start of target trials to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the msc sequence, $V_{TA}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the dev sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.

Next, we turned to the likely cause of this difference, the non-target responses (Figure \ref{fig5}b), mapped into space as the average number of spikes fired in each neuron. As expected, the pattern of activity in non-target trials, constituting 80\% of all trials, closely matched that of the thresholds. Notably, although the comparison is between 400 B trials in dev, and only 100 B trials (alongside the same number of C, D and E trials) in msc, only a small handful of neurons responded more to non-target stimulation in dev, most of these very close to the stimulation site of B.

Correlating the non-target response contrast (dev - msc) with the contrast in $V_{TA}$ across excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev almost perfectly predicted the resulting reduction in TA levels (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response between msc and dev sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the non-target trial response is the primary driver of TA.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any condition. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (dev - msc) and the contrast in $V_{TA}$ at the start of B trials, and associated linear regression. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are plotted with crosses, while later responders are plotted with dots. Inset histograms show the distributions of the corresponding values across these two categories, and the confidence intervals of their medians, calculated by bootstrap with 10000 iterations.
\textbf{c} Median non-target response contrast (dev - msc) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labelled "late", across networks and stimuli.
\textbf{c} Median contrast (dev - msc) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and thereby decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (1) responds very quickly and comparatively strongly to stimulation in B, and therefore more to non-target stimuli in dev than msc, and (2) has increased TA levels as a result, reducing the response of the remaining downstream neurons in the same way as demonstrated in reverse in Figure \ref{fig4}.

To see this in the sample network, consider first the latency of responses to B across all conditions, shown by way of rank order in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all conditions separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B in any condition, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast (dev - msc) to the contrast in $V_{TA}$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993) echoed the related finding in Figure \ref{fig5}c. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the dev sequence, while the response of the network as a whole decreased. Correspondingly, the TA levels of this portion of the network also tended to increase, in clear contrast to the network as a whole.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $V_{TA}$ contrasts were robustly positively correlated (median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $V_{TA}$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in dev than msc (T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $V_{TA}$ followed the same pattern, with the early portion exhibiting stronger TA in dev (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).
